
---
title: RR WEEK 2 - COURSE PROJECT 1
output: 
  html_document:
    keep_md: true


---


# 1. Loading and preprocessing data 

Here is the code for reading in and preprocessing the data. 

```{r}

library(ggplot2) # load required packages
    
data <- read.csv("activity.csv") # read in data 
data_clean <- data[!is.na(data$steps),] # remove rows with NA values i.e. consider only clean/tidy data  

```

```{r global_options}
knitr::opts_chunk$set(fig.path='Figs/')
```

# 2. Histogram of the total number of steps taken each day.  

For the clean data the total number of steps is calculated and plotted in a histogram. 

```{r}

sum_steps <- aggregate(data_clean$steps, by = list(data_clean$date), sum) # sum steps of clean data per date 
names(sum_steps) <- c("date","steps") # rename columns of result
    
hist_steps <- ggplot(sum_steps, aes(date,steps)) + geom_histogram(stat = "identity")+ ggtitle("Number of steps per day")  # create histogram
print(hist_steps)
    
```

# 3. Mean and median of total number of steps   

Furthermore, mean and median of the total number of steps is calculated. 

```{r}

mean(sum_steps$steps) # calculate mean
median(sum_steps$steps) # calculate median 
    
```

# 4. Time series plot of the average number of steps taken

Next, the average number of steps is calculated per interval and plotted in a time series. 

```{r}

mean_int <- aggregate(data_clean$steps,by = list(data_clean$interval), mean) # calculate average number of steps per interval
names(mean_int) <- c("interval","mean(steps)") # rename columns of result 
    
plot(mean_int$interval, mean_int$`mean(steps)`, type = "l", xlab = "Interval", ylab = "Avg(# steps)", 
    main = "Average number of steps per interval", col = "blue") # create time series 
    
```

# 5. The 5-minute interval that, on average, contains the maximum number of steps

This interval is calculated as follows:

```{r}

mean_int[mean_int[,2] == max(mean_int$`mean(steps)`), 1] # determine interval with maximum numver of steps 
 
```

# 6. Code to describe and show a strategy for imputing missing data

The total number of missing values is: 

```{r}

sum(!is.na(data$steps)) # go back to raw data (variable data) and determine # NAs
 
```

Now we want to fill in the missing values. Each NA value is set to be the average number of steps for the corresponding interval for the tidy data.    

```{r}

m_data <- merge(data,mean_int, by.x = "interval", by.y = "interval", all.x = TRUE) # merge raw data (data) and table with avg # of steps per interval (mean_int)
    
for (i in which(is.na(m_data[,2]))) { # for each NA set the missing value to be the avg # of steps of the corresponding interval 
    m_data[i,2] <- m_data[i,4]
} 
    
m_data$`mean(steps)` <- NULL # to get the data with imputed missing values in the format format like the data with missing values delete last column
m_data <- m_data[,c("steps", "date","interval")] # and reorder columns
    
sum_steps_m <- aggregate(m_data$steps, by = list(m_data$date), sum) # calculate # steps per day with imputed missing values 
names(sum_steps_m) <- c("date","steps") # rename result 
    
 
```


# 7. Histogram of the total number of steps taken each day after missing values are imputed

The histogram for the dataset with imputed missing values is: 

```{r}
hist_steps_NA_imputed <- ggplot(sum_steps_m, aes(date,steps)) + geom_histogram(stat = "identity")+ ggtitle("Number of steps per day with missing values imputed") 

print(hist_steps_NA_imputed)
 
```

Mean and median are:

```{r}

mean(sum_steps_m$steps) # calculate mean
median(sum_steps_m$steps) # calculate median
 
```

# 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}

m_data$dayType <- ifelse(weekdays(as.Date(m_data$date)) %in% c("Samstag","Sonntag"), "weekend", "weekday") # create factor variable indicating whether day is weekday or weekend 
ggplot(m_data, aes(interval,steps)) + stat_summary(fun.y = mean, geom = "line", colour = "blue") + facet_wrap(~dayType) + ggtitle("Average number of steps per interval") # plot average number of steps per interval, seperately for weekdays and weekend 
 
```

# 9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

Number 1-8 contain the respective code chunks to reproduce the results.  
